Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex
Added machine info

\begin{abstract}
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs.  To do this, we  analyze programs for
liveness of  heap-allocated data.  The accompanying  garbage collector
is modified  to use the result  of the analysis to  preserve only live
data---a subset  of reachable data---during garbage  collection.  This
results  in an  increase  in the  garbage  reclaimed and  consequently
reduces the  memory footprint of  programs.  While this  technique has
already been shown to yield  benefits for eager first-order languages,
the lack of a statically determinable execution order and the presence
of closures at run-time pose additional challenges for lazy languages.
This requires changes  both in the earlier liveness  analysis and also
in the  garbage collection scheme  required to reclaim memory  that is
not live.

@@

The specific contributions of this work are
  \begin{itemize}
  \item  We formulate  a  liveness  analysis for  a  lazy first  order
    functional language. We show how  to compute function summaries in
    a  context  independent  manner  to achievs  a  context  sensitive
    analysis. We  further describe the transformation  of the liveness
    equations to grammar rules in a  way that the language accepted by
    the resulting grammar describes live heap locations.\mycomment{the
      proof of its correctness}
  \item We prove  the undecidability of the grammar  that results from
    the  analysis.   To overcome  this,  we  present an  algorithm  to
    transform grammar  to finite  state automata  (DFA) such  that the
    language accepted by the  DFA over-approximates the liveness (and
    hence a safe approximation).
  \item We have implemented a garbage  collector that can use the DFA
    to  retain live  cells that  also include  closures. We  have done
    extensive  experiments with  this garbage  collector and  obtained
    empirical results  that show  the effectiveness  of liveness-based
    garbage collection.
  \end{itemize}

@@

Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
language  considered  in our  analysis  and  a small-step  operational
semnatics  for it.   We present  our liveness  analysis for  this lazy
language  in Section~\ref{sec:liveness}.   Section~\ref{sec:computing}
describes  the  formulation of  the  liveness  as  a grammar  and  its
approximation  to   DFA.  It  also   contains  a  proof   showing  the
undecidability  of  the  precise  liveness in  our  formulation  which
necessitates  the approximation  to DFA.   Section~\ref{sec:GC-scheme}
discusses different ways of handling closures and their pros and cons.
We  report our  experimental results  in Section~\ref{sec:experiments}
along with some observations.  Section~\ref{sec:relatedwork} discusses
previous  work   related  to  garbage  collection   and  liveness  and
Section~\ref{sec:conclusion}   discusses   possible   extensions   and
concludes the paper.
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective} where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of  closures through  the $\LET$  construct.  $\LET$s  in our
language  are lazy;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
\IN\,\, e$, $x$ may occur in  $s$. This enables creation of graph-like
structures in  a pure language. The  restriction of \LET\ to  a single
definition  is for  ease  of  exposition---generalization to  multiple
definitions does not add conceptual difficulties.  We further restrict
each variable in a program to  be distinct, so that no scope shadowing
occurs---this  simplifies  the  reasoning   about  the  program.   The
$\SRETURN$  and   \SIF\  expressions  and  strict   operators  trigger
evaluation of closures.

@@

\begin{figure*}[t!]
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

@@

The  semantics  of   expressions  (and  applications\footnote{In  most
  contexts, we  shall use  the term 'expression' and the notation $e$  for
both  expressions and
  applications.}) are  given by transitions  of the form  $\rho, \stk,
\heap, e \rightarrow \rho', \stk', \heap', e'$.  Here \stk\ is a stack
of  continuation  frames.  Each  continuation  frame  is  of the  form
$(\rho, x, e)$, signifying that the  location of $x$ has to be updated
with the  value of the currently  evaluating expression, and $e$  is to be
evaluated  next  in  the  environment  $\rho$.   The  start  state  is
$([\;]_\rho,([\;]_\rho,  \ans,   (\print~\ans)):[\;]_{S}  ,  [\;]_{H},
(\mainpgm)$.  In  this, the initial  environment and the  initial heap
are  denoted  by  $[\;]_\rho$  and  $[\;]_H$, and  the  initial  stack
consists  of  a  single  continuation   frame  in  which  \ans\  is  a
distinguished  variable  that  would  be  updated with  the  value  of
(\mainpgm).  In  addition, \print\ is a function  modelling a printing
mechanism---a standard  runtime support assumption  for lazy languages
that prints the  value of (\mainpgm).  Notice the  use of the operator
$:$ for adding elements to the top of the stack.

%% In absence of a static type-checking mechanism, programs with other
%% than syntactic  errors lead to states  not covered by  any rules in
%% the small-step semantics.
The notation  $[\myvec{x}   \mapsto  \myvec{\ell}]$   represents  an
environment that maps variables $x_i$
%$x_1,  \ldots, x_n$
to locations $\ell_i$
%$\ell_1,  \ldots,  \ell_n$,
and  $\heap[\ell :=  d]$
indicates the  updation of  a heap \heap\  at $\ell$ with  $d$.  $\rho
\oplus \rho'$  represents the  environment $\rho$ shadowed  by $\rho'$
and $\lfloor \rho \rfloor_X$  represents the environment restricted to
the locations in $X$. Finally $FV(s)$ represents the free variables in
the application $s$.

@@

We thus  regard the  body of  the function as  an expression  tree and
consider paths from  the root of this tree to  the nodes corresponding
to the program points which trigger evaluation\footnote{Program points
  which      trigger      evaluation       are      labelled      with
  $\psi$.}---\SIF\ conditions and \SRETURN\ expressions.  We call such
paths {\em e-paths}.  For example, for the function \length, there are
three  such paths  $\epath_1 =  \langle \pi_1,  \pi_2, \psi_1\rangle$,
$\epath_2  = \langle\pi_1,  \pi_2,  \pi_3,  \pi_4, \psi_2\rangle$  and
$\epath_3  =  \langle  \pi_1,   \pi_2,  \pi_5,  \pi_6,  \pi_7,  \pi_8,
\psi_3\rangle$.

@@

Consider the  $\mathcal{L}$-rules for {\LET}, {\SIF},  and {\SRETURN}.
The expression  $(\SRETURN~\psi:x)$ with  a demand $\sigma$  forces an
evaluation  of  $x$ and  hence  a  new map  $[\mathit{\epath}  \mapsto
  \{x.\sigma\}]  $ needs  to be  created,  where $\epath$  is a  fresh
evaluation path  terminating in  $\psi$.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union of  the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$, the map $[\mathit{\epath} \mapsto \{x.\epsilon\}]$ is created and
added  to the  union.  Also  notice a  consequences of  laziness---the
entire  expression including  the condition  is not  evaluated if  the
demand on it is $\emptyset$.  This  results in the empty map, which is
denoted by $\emptyset_{\cal{M}}$.

@@

Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the grammar for \var{\Df{\length}{1}}.  The final DFA is shown in
Figure~\ref{fig:example-automata}(b).  This expectedly says that given
a  demand $\epsilon$  on  \length,  the liveness  of  its argument  is
$\acdr^{*}$  (the  spine  of   the  list  is  traversed).   Similarly,
Figure~\ref{fig:example-automata}(c)     shows     the     NFA     for
\var{\Lanv{9}{\pa}}.          The         DFA         shown         in
Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
paths.   This  reflects  the  lazy  nature  of  our  language.   Since
\length\ does not  evaluate the member elements of  the argument list,
the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
liveness based GC is triggered beyond $\pi_9$.

@@

\begin{table*}[ht!]
\caption{Comparing LGC times for different strategies}
\label{tab:compare-live-clo-reach-clo}
\centering
\begin{tabular}
{| l | r | r | r | r |}
\hline
Prog Name & Heap size & Num GC & Time taken using reachablity for closure  & Time taken using liveness for closure \\ \hline
\verb@nqueens(8)@ & 10000 & 183 & 0.0087s & 0.2267s\\ \hline
\verb@lambda@&10000&11&0.0010s&4.9922s\\ \hline
\verb@lcss@&10000&1&0.00008s&34.6684s\\ \hline
\end{tabular}
\end{table*}
To compare the  effect of using reachability to  copy closures instead
of liveness we take some representative programs from our benchmark suite and
execute       them      with       both.        As      shown       in
Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
collections in both cases are same but copying closures using liveness
takes a lot more time compared to reachability based copying.

@@

Table~\ref{tab:experimental-results} shows  that  LGC
improves the  garbage collection time  and efficiency for most  of the
programs.  Even for  programs where the time taken is  more it is very
competitive.  Memory usage  graphs are shown for  select benchmarks in
Figure~\ref{fig:memory-usage}.  As  the number of  garbage collections
tend  to  be  very large  our  graphs  show  only  a window  which  is
representative of the behavior for  that particular benchmark.  In all
the programs we can see that  the curve corresponding to LGC regularly
dips below the RGC curve.

One area  of concern is the  huge gap between the  actual liveness and
the liveness  percieved by  our collector.  In case  of LGC  for eager
languages  the gap  was  very  narrow and  almost  touched the  actual
liveness curve.  In case of  lazy languages, due to using reachability
for copying we  end up copying more  cells. Any data which  is part of
closure irrespective  of whether it is  live or not gets  copied. Thus
many non-live  cells get  copied during garbage  collection increasing
the drag  time.  Our  experiments have  shown that  unless we  use the
exact  execution  point  for  closures, bringing  down  the  number  of
non-live  cells  copied in  closures  cannot  be reduced.   Using  the
approximated liveness (liveness  at the creation point)  does not give
benefits  in terms  of non-live  cells inside  closures as  everything
would be live at the point of creation.

@@

Another  interesting  research  in  this   field  is  the  concept  of
Simplifiers~\cite{ONeill}. They are  described as lightweight daeomons
which improve the efficiency of the program in general and the garbage
collector in particular. Most of the simplifications mentioned in this
work are subsumed by a liveness  based collector. We feel this work is
orthogonal and can augment our  liveness based collector.  The closest
to our  approach is the  liveness based garbage  collector implemented
in~\cite{asati14lgc}. We  extend their work to  handle lazy evaluation
and closures.

@@

%% One issue we  highlighted was that while fewer  nodes were marked (and
%% more garbage collected), sometimes  cons cells could be visited and
%% traversed  multiple times  with different  sets of  liveness  paths to
%% explore.  Further work includes  improvements to the classical copying
%% collector to reduce the cost of this.
We extended the liveness based garbage collection to lazy
  languages and shown its benefit  for garbage collection in practical
  programs. We  defined context  senstive liveness analysis  that uses
  context independent summaries of functions obtained using a symbolic
  demand. We show  that obtaining precise solution  to these equations
  is undecidable in  our formulation, and hence  we safely approximate
  the result using DFA. These DFA are consulted by the garbage collector
  to improve  collection.

>>>>

\begin{abstract}
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs.  To do this, we  analyze programs for
liveness of  heap-allocated data.  The accompanying  garbage collector
is modified  to use the result  of the analysis to  preserve only live
data---a subset  of reachable data---during garbage  collection.  This
results  in an  increase  in the  garbage  reclaimed and  consequently
reduces the  memory footprint of  programs.  While this  technique has
already been shown to yield  benefits for eager first-order languages,
the lack of a statically determinable execution order and the presence
of closures at runtime pose additional challenges for lazy languages.
This requires changes  both in the earlier liveness  analysis and also
in the  garbage collection scheme  required to reclaim memory  that is
not live.

@@

The specific contributions of this work are
  \begin{itemize}
  \item  We formulate  a  liveness  analysis for  a  lazy first  order
    functional language. We show how  to compute function summaries in
    a  context  independent  manner  to achieve  a  context  sensitive
    analysis. We  further describe the transformation  of the liveness
    equations to grammar rules in a  way that the language accepted by
    the resulting grammar describes live heap locations.\mycomment{the
      proof of its correctness}
  \item We prove  the undecidability of the grammar  that results from
    the  analysis.   To overcome  this,  we  present an  algorithm  to
    transform grammar  to finite  state automata  (DFA) such  that the
    language accepted by the  DFA over-approximates the liveness (and
    hence a safe approximation).
  \item We have implemented a garbage  collector that can use the DFA
    to  retain live  cells that  also include  closures. We  have done
    extensive  experiments with  this garbage  collector and  obtained
    empirical results  that show  the effectiveness  of liveness-based
    garbage collection.
  \end{itemize}

@@

Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
language  considered  in our  analysis  and  a small-step  operational
semantics  for it.   We present  our liveness  analysis for  this lazy
language  in Section~\ref{sec:liveness}.   Section~\ref{sec:computing}
describes  the  formulation of  the  liveness  as  a grammar  and  its
approximation  to   DFA.  It  also   contains  a  proof   showing  the
undecidability  of  the  precise  liveness in  our  formulation  which
necessitates  the approximation  to DFA.   Section~\ref{sec:GC-scheme}
discusses different ways of handling closures and their pros and cons.
We  report our  experimental results  in Section~\ref{sec:experiments}
along with some observations.  Section~\ref{sec:relatedwork} discusses
previous  work   related  to  garbage  collection   and  liveness  and
Section~\ref{sec:conclusion}   discusses   possible   extensions   and
concludes the paper.
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective} where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of  closures through  the $\LET$  construct.  $\LET$s  in our
language  are lazy;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
\IN\,\, e$, $x$ may occur in  $s$. This enables creation of graph-like
structures in  a pure language. The  restriction of \LET\ to  a single
definition  is for  ease  of  exposition---generalization to  multiple
definitions does not add conceptual difficulties.  We further restrict
each variable in a program to  be distinct, so that no scope shadowing
occurs---this  simplifies  the  reasoning   about  the  program.   The
$\SRETURN$  and   \SIF\  expressions  and  strict   operators  trigger
evaluation of closures.

@@

\begin{figure*}[t!]
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[\rho'(z) := (\rho(x), \rho(y))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

@@

The  semantics  of   expressions  (and  applications\footnote{In  most
  contexts, we  shall use  the term 'expression' and the notation $e$  for
both  expressions and
  applications.}) are  given by transitions  of the form  $\rho, \stk,
\heap, e \rightarrow \rho', \stk', \heap', e'$.  Here \stk\ is a stack
of  continuation  frames.  Each  continuation  frame  is  of the  form
$(\rho, x, e)$, signifying that the  location of $x$ has to be updated
with the  value of the currently  evaluating expression, and $e$  is to be
evaluated  next  in  the  environment  $\rho$.   The  start  state  is
$([\;]_\rho,([\;]_\rho,  \ans,   (\print~\ans)):[\;]_{S}  ,  [\;]_{H},
(\mainpgm)$.  In  this, the initial  environment and the  initial heap
are  denoted  by  $[\;]_\rho$  and  $[\;]_H$, and  the  initial  stack
consists  of  a  single  continuation   frame  in  which  \ans\  is  a
distinguished  variable  that  would  be  updated with  the  value  of
(\mainpgm).  In  addition, \print\ is a function  modeling a printing
mechanism---a standard  run-time support assumption  for lazy languages
that prints the  value of (\mainpgm).  Notice the  use of the operator
$:$ for adding elements to the top of the stack.

%% In absence of a static type-checking mechanism, programs with other
%% than syntactic  errors lead to states  not covered by  any rules in
%% the small-step semantics.
The notation  $[\myvec{x}   \mapsto  \myvec{\ell}]$   represents  an
environment that maps variables $x_i$
%$x_1,  \ldots, x_n$
to locations $\ell_i$
%$\ell_1,  \ldots,  \ell_n$,
and  $\heap[\ell :=  d]$
indicates the  updating of  a heap \heap\  at $\ell$ with  $d$.  $\rho
\oplus \rho'$  represents the  environment $\rho$ shadowed  by $\rho'$
and $\lfloor \rho \rfloor_X$  represents the environment restricted to
the locations in $X$. Finally $FV(s)$ represents the free variables in
the application $s$.

@@

We thus  regard the  body of  the function as  an expression  tree and
consider paths from  the root of this tree to  the nodes corresponding
to the program points which trigger evaluation\footnote{Program points
  which      trigger      evaluation       are      labeled      with
  $\psi$.}---\SIF\ conditions and \SRETURN\ expressions.  We call such
paths {\em e-paths}.  For example, for the function \length, there are
three  such paths  $\epath_1 =  \langle \pi_1,  \pi_2, \psi_1\rangle$,
$\epath_2  = \langle\pi_1,  \pi_2,  \pi_3,  \pi_4, \psi_2\rangle$  and
$\epath_3  =  \langle  \pi_1,   \pi_2,  \pi_5,  \pi_6,  \pi_7,  \pi_8,
\psi_3\rangle$.

@@

Consider the  $\mathcal{L}$-rules for {\LET}, {\SIF},  and {\SRETURN}.
The expression  $(\SRETURN~\psi:x)$ with  a demand $\sigma$  forces an
evaluation  of  $x$ and  hence  a  new map  $[\mathit{\epath}  \mapsto
  \{x.\sigma\}]  $ needs  to be  created,  where $\epath$  is a  fresh
evaluation path  terminating in  $\psi$.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a point-wise-union of  the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$, the map $[\mathit{\epath} \mapsto \{x.\epsilon\}]$ is created and
added  to the  union.  Also  notice a  consequences of  laziness---the
entire  expression including  the condition  is not  evaluated if  the
demand on it is $\emptyset$.  This  results in the empty map, which is
denoted by $\emptyset_{\cal{M}}$.

@@

\subsubsection{From CFGs to GC-ready DFA}
\label{sec:NFA-approx}
 We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
 approximate a CFG to a {\em strongly regular\/} grammar.  For
 example, the grammar
 fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
 Mohri-Nederhof transformation is:
 \begin{eqnarray*}
   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
   \acdr\var{\Df{\length}{1}}
   \mid \clazy\var{\Df{\length}{1}}\\
   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
   \mid \epsilon
 \end{eqnarray*}

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges labeled  \bcar\acar\ or  \bcdr\acdr\ and
constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
simplification is continued till a fixed point is reached, after which
the edges labeled \bcar \ and  \bcdr are deleted.  The details of the
algorithm, its correctness and termination proofs are given by Karkare
et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
has  edges labeled  with  \acar,  \acdr\ and  \clazy\  only. In  this
automaton, for every edge labeled \clazy, we check if the source node
of the edge has a valid path to  a final final state.  If yes, we mark
the mark  the source node as  final. Finally, we remove  all the edges
labeled  \clazy\  and  convert  the automaton  into  a  deterministic
automaton.    This   effectively  implements   the   $\hookrightarrow$
simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
paths. While  checking for liveness  during GC, a  forward access-path
only if it can reach a final state.
Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the grammar for \var{\Df{\length}{1}}.  The final DFA is shown in
Figure~\ref{fig:example-automata}(b).  This expectedly says that given
a  demand $\epsilon$  on  \length,  the liveness  of  its argument  is
$\acdr^{*}$  (the  spine  of   the  list  is  traversed).   Similarly,
Figure~\ref{fig:example-automata}(c)     shows     the     NFA     for
\var{\Lanv{9}{\pa}}.          The         DFA         shown         in
Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
paths.   This  reflects  the  lazy  nature  of  our  language.   Since
\length\ does not  evaluate the member elements of  the argument list,
the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
liveness based GC is triggered beyond $\pi_9$.

@@

\begin{table*}[ht!]
\caption{Comparing LGC times for different strategies}
\label{tab:compare-live-clo-reach-clo}
\centering
\begin{tabular}
{| l | r | r | r | r |}
\hline
Prog Name & Heap size & Num GC & Time taken using reachability for closure  & Time taken using liveness for closure \\ \hline
\verb@nqueens(8)@ & 10000 & 183 & 0.0087s & 0.2267s\\ \hline
\verb@lambda@&10000&11&0.0010s&4.9922s\\ \hline
\verb@lcss@&10000&1&0.00008s&34.6684s\\ \hline
\end{tabular}
\end{table*}
To compare the  effect of using reachability to  copy closures instead
of liveness we take some representative programs from our benchmark suite and
execute       them      with       both.        As      shown       in
Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
collections in both cases are same but copying closures using liveness
takes a lot more time compared to reachability based copying.

@@

Table~\ref{tab:experimental-results} shows  that  LGC
improves the  garbage collection time  and efficiency for most  of the
programs.  Even for  programs where the time taken is  more it is very
competitive.  Memory usage  graphs are shown for  select benchmarks in
Figure~\ref{fig:memory-usage}.  As  the number of  garbage collections
tend  to  be  very large  our  graphs  show  only  a window  which  is
representative of the behavior for  that particular benchmark.  In all
the programs we can see that  the curve corresponding to LGC regularly
dips below the RGC curve. All our benchmarks were executed on a desktop with 64 bit Ubuntu 14.04,  8 core
Intel(R) Core(TM) i7-4770 CPU @ 3.40GHz, L2 cache of 8192 KB and 
15922 MiB (16 GB) RAM.

One area  of concern is the  huge gap between the  actual liveness and
the liveness  perceived by  our collector.  In case  of LGC  for eager
languages  the gap  was  very  narrow and  almost  touched the  actual
liveness curve.  In case of  lazy languages, due to using reachability
for copying we  end up copying more  cells. Any data which  is part of
closure irrespective  of whether it is  live or not gets  copied. Thus
many non-live  cells get  copied during garbage  collection increasing
the drag  time.  Our  experiments have  shown that  unless we  use the
exact  execution  point  for  closures, bringing  down  the  number  of
non-live  cells  copied in  closures  cannot  be reduced.   Using  the
approximated liveness (liveness  at the creation point)  does not give
benefits  in terms  of non-live  cells inside  closures as  everything
would be live at the point of creation.

@@

Another  interesting  research  in  this   field  is  the  concept  of
Simplifiers~\cite{ONeill}. They are  described as lightweight daemons
which improve the efficiency of the program in general and the garbage
collector in particular. Most of the simplifications mentioned in this
work are subsumed by a liveness  based collector. We feel this work is
orthogonal and can augment our  liveness based collector.  The closest
to our  approach is the  liveness based garbage  collector implemented
in~\cite{asati14lgc}. We  extend their work to  handle lazy evaluation
and closures.

@@

%% One issue we  highlighted was that while fewer  nodes were marked (and
%% more garbage collected), sometimes  cons cells could be visited and
%% traversed  multiple times  with different  sets of  liveness  paths to
%% explore.  Further work includes  improvements to the classical copying
%% collector to reduce the cost of this.
We extended the liveness based garbage collection to lazy
  languages and shown its benefit  for garbage collection in practical
  programs. We  defined context  sensitive liveness analysis  that uses
  context independent summaries of functions obtained using a symbolic
  demand. We show  that obtaining precise solution  to these equations
  is undecidable in  our formulation, and hence  we safely approximate
  the result using DFA. These DFA are consulted by the garbage collector
  to improve  collection.
